---
title: "Exploratory analysis script"
author: "Chaohua Li"
date: "02/23/2024"
output: html_document
---

# Load packages

```{r}
library(here)  
library(dplyr)
library(skimr)
library(ggplot2)
library(gtsummary)
library(knitr)
```

# Load the clean data.

```{r}
data_location <- here("data","processed-data","processeddata.rds")
project <- readRDS(data_location)
```


# Data exploration 

Explore numeric variables one by one using tables and histograms.

```{r}
#get descriptive statistics of variables and create a table
summary_df<-gtsummary::tbl_summary(select(project,c(-"TL_GEO_ID",-"state")), statistic = list(
  all_continuous() ~ "{mean}/{sd}",
  all_categorical() ~ "{n} / {N} ({p}%)"
),)
knitr::kable(summary_df,caption = "Summary Table: Mean/SD or n/N(%)")

# save table in results folder
summarytable_file = here("results","exploratory", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)


#Histograms to inspect distribution
p1<-hist(project$Rate2 )
p2<-hist(project$pcp_100k )
p3<-hist(project$hispanic_pct )
p4<-hist(project$NHB_pct )
p5<-hist(project$noHS_pct )
p6<-hist(project$poverty_pct )
p7<-hist(project$unemployed_pct )
p8<-hist(project$uninsured_pct )

#Save all the histograms in a 8-panel png file
par(mfrow = c(2, 4))
plot(p1)
plot(p2)
plot(p3)
plot(p4)
plot(p5)
plot(p6)
plot(p7)
plot(p8)
par(mfrow = c(1, 1)) # Reset the layout
histogram_file = here("results","exploratory", "combined_histogram.png")
dev.copy(png, histogram_file) # Save the combined plot
dev.off()  

```
From the histograms, we can tell the distribution for the outcome (mortality rate) is 
fairly normal. Distributions for PCP supply, %Hispanic and %NH-Black are very skewed.
We probably have to log-transform PCP supply. 


# Scatterplot: Mortality Rate vs log(PCP supply)

```{r}
#examine the association between Mortality Rate and log(PCP supply)
Rate_PCP1 <- project %>% ggplot(aes(x=log(pcp_100k), y=Rate2)) + geom_point() + geom_smooth(method='lm') 
plot(Rate_PCP1 )
figure_file1 = here("results","exploratory","rate_pcp1.png")
ggsave(filename = figure_file1, plot=Rate_PCP1) 
```
It seems there is a linear relationship between Mortality Rate and log(PCP supply). 
Then I would like to inspect the relationship by Metro status of the county. 

```{r}
Rate_PCP2 <- project %>% ggplot(aes(x=log(pcp_100k), y=Rate2)) + geom_point() + geom_smooth(method='lm')+facet_wrap(~ metro, scales = "free_y") 
plot(Rate_PCP2 )
figure_file2 = here("results","exploratory","rate_pcp2.png")
ggsave(filename = figure_file2, plot=Rate_PCP2) 
```

The relationship looks similar for the metro and non-metro counties. Maybe there is 
no interaction by metro status. 


# Inspect correlation among covariates

```{r}
#create correlation matrix for the 6 covariates
cor<-cor(project[,c(-1,-2,-3,-4,-5)])
knitr::kable(cor,caption = "Correlation among covariates")
# save table in results folder
cor_path = here("results","exploratory", "correlation_table.rds")
saveRDS(cor, file = cor_path)
```

The correlations among the covariates are relatively low. We can keep all the covariates 
for modeling. 




